What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Expand conclusion

\paragraph{Consistency} It is interesting to consider how many different
forms any given site sees. The more forms, the harder it will be to characterize
the behavior of the program at that site. Luckily, \packageNbOneMinimizedPercent
of sites only see a single form. There are few sites that are highly polymorphic
(8 or more different forms), these include the pipe operator of the \c{magrittr}
package which is used to compose functions.

@@

\noindent
As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
function environment with the \c{global} kind as a distant second. This means
that most variable lookups and most side effects either read/update existing
local variables or introduce new ones. But for which function? From the point of
view of the function that called \eval, Table~\ref{tab:funoffset} gives an
offset on the call stack. Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on. The data suggests that in 81\% of cases, \eval accesses its caller's environment -- this means, the variables of the function where \eval textually
occurs are read and written to. It is interesting that some 1.5\% of sites
evaluate a code in an environment that is three frames or above from \eval. This
distance implies that, in general, modular reasoning is impossible, to
understand the behavior of any piece of code entails fully understanding the
behavior of all functions that the code may call as the actions of \eval may
happen at a distance. Finally note that any particular site may have several
kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
site has a single kind. It means that we can predict the \emph{shape} of the expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites.

@@

In this paper, we provide a large-scale study of the use of the \eval function
in the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries. We have found just a few \eval
call sites in scripts coming from large corpus of Kaggle that represents code
written by R practitioners. We used an instrumented version of the R virtual
machine, \rdyntrace, to capture all \eval calls traces including the relevant
side effects that happen within these calls. Several insights can be learnt
from the data.
%
First, \eval is widely used in the R core packages that come bundled with the
language. Because of that, there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere which is wary for both R developers and R compiler.
% TODO expressions passed to eval
Even more so in the case of side effects that can quite literally happen
anywhere. The data however suggest that this is not the case. The majority of
eval sites do not do any observable side effects and when they do, it happens
in a predictable environment.

% how much can we say about the threat model?
Similarly to \citep{ecoop11}, we started looking into \evals in R with the
ambition to replace them by other feature that are more amenable to static
analysis. The data unfortunately suggests that replacing \eval is going to be
lengthly journey. Even if we disregard core packages as they are part of the
language and the \eval use is stable, in CRAN \eval is used widely and in more
sophisticated way than was the case in JavaScript. There the vast majority of
sites fitted an extremely simple categorization based on simple regular
expressions. Not so in R. While some usage, especially the one which calls
\eval with an expression parsed from string, can be categorized and often
replaced by equivalent yet more disciplined and safer code, the case of \eval
coming from reflection is much harder.

>>>>

\paragraph{Consistency} It is interesting to consider how many different
forms any given site sees. The more forms, the harder it will be to characterize
the behavior of the program at that site. Luckily, \packageNbOneMinimizedPercent
of sites only see a single form.  In other words, a JIT compiler could correctly predict the \emph{shape} of an expression by looking just at the first call in \packageNbOneMinimizedPercent of the sites.  There are few sites that are highly polymorphic (8 or more different forms), these include the pipe operator of the \c{magrittr} package which is used to compose functions.

@@

\noindent
As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
function environment with the \c{global} kind as a distant second. This means
that most variable lookups and most side effects either read/update existing
local variables or introduce new ones. But for which function? From the point of
view of the function that called \eval, Table~\ref{tab:funoffset} gives an
offset on the call stack. Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on. The data suggests that in 81\% of cases, \eval accesses its caller's environment -- this means, the variables of the function where \eval textually
occurs are read and written to. It is interesting that some 1.5\% of sites
evaluate a code in an environment that is three frames or above from \eval. This
distance implies that, in general, modular reasoning is impossible, to
understand the behavior of any piece of code entails fully understanding the
behavior of all functions that the code may call as the actions of \eval may
happen at a distance. Finally note that any particular site may have several
kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
site has a single kind. It means that we can predict the \emph{kind} of the expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites.

@@

In this paper, we provide a large-scale study of the use of the \eval function
in the R programming language. It is based primarily on a corpus of
\CranRunnableScripts scripts extracted from \CranPackages R packages. As
expected, \eval is used mainly in libraries. We have found just a few \eval
call sites in scripts coming from large corpus of Kaggle that represents code
written by R practitioners. We used an instrumented version of the R virtual
machine, \rdyntrace, to capture all \eval calls traces including the relevant
side effects that happen within these calls. Several insights can be learnt
from the data.
%
First, \eval is widely used in the R core packages that come bundled with the
language. Because of that, there is hardly any R code that would not use it.
\eval is used to implement some basic functionality such as package loading.
Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
the available \CranPackages packages.
%
The code executed by R \eval has unusual expressive power. It can essentially
modify anything anywhere which is wary for both R developers and R compiler. The shapes of expressions passed to \eval are extremely diverse, ranging from simple variable lookups to complex blocks or creation of new functions. Luckily, most site only see a single shape. \eval unlashes its power and shows up its singularity compared to \eval in other languages with the diversity of environments in which it can evaluate its expressions. While most of the time, the \eval environment is the local environment, it is also often the global environment, a synthetic environment, or even a package environment. Luckily, as for the expressions passed to \eval, the environment passed to \eval for one given site is easily predictable with the first call to the site.

The expressive power of \eval shines even more  so in the case of side effects that can quite literally happen anywhere. The data however suggest that this is not the case. The majority of eval sites do not do any observable side effects and when they do, it happens in a predictable environment.

The origins of the expression passed to \eval, either constructed from language expressions, or stemming from reflection, or built by parsing strings. Reflection and constructed expressions dominate, which suggest the main use of \eval is meta-programming. % macros...

% how much can we say about the threat model?
Similarly to \citep{ecoop11}, we started looking into \evals in R with the
ambition to replace them by other feature that are more amenable to static
analysis. The data unfortunately suggests that replacing \eval is going to be
a lengthy journey. Even if we disregard core packages as they are part of the
language and the \eval use is stable, in CRAN \eval is used widely and in more
sophisticated way than was the case in JavaScript. There the vast majority of
sites fitted an extremely simple categorization based on simple regular
expressions. Not so in R. While some usage, especially the one which calls
\eval with an expression parsed from string, can be categorized and often
replaced by equivalent yet more disciplined and safer code, the case of \eval
coming from reflection is much harder.
